---
title: "Diagnosing Dynamic Data Step by Step"
author: "Ivan Jacob Agaloos Pesigan"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{Diagnosing Dynamic Data Step by Step}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r}
#| include = FALSE
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>"
)
```

This vignette shows a transparent preprocessing and diagnostic workflow for
intensive longitudinal or dynamic modeling data. The goal is to make each
decision inspectable rather than hiding all decisions inside one large
preprocessing wrapper.

```{r}
library(dynTools)

data <- data.frame(
  id = c(1, 1, 1, 2, 2, 2, 3, 3),
  time = c(0, 1, 3, 0, 1, 2, 0, 1),
  y1 = c(1, 2, 3, NA, NA, 4, Inf, 2),
  y2 = c(2, NA, 4, NA, NA, 5, 1, NaN),
  cov = c(10, 11, 13, 20, 21, 22, 30, 31)
)

id <- "id"
time <- "time"
observed <- c("y1", "y2")
covariates <- "cov"
```

## 1. Check structure

```{r}
CheckDynData(
  data = data,
  id = id,
  time = time,
  observed = observed,
  covariates = covariates,
  require_unique = FALSE,
  require_numeric_time = TRUE,
  require_numeric_observed = TRUE,
  require_numeric_covariates = FALSE,
  min_rows = 1L
)
```

## 2. Visualize trajectories by ID

A visual check is a useful first diagnostic step. `PlotByID()` overlays the
observed trajectories by ID. Non-finite observed values such as `NaN`, `Inf`,
and `-Inf` are treated as missing for plotting, but they should still be
addressed explicitly in the numerical diagnostics.

```{r plot-raw, fig.width = 7, fig.height = 4}
PlotByID(
  data = data,
  id = id,
  time = time,
  observed = observed,
  legend = TRUE,
  ask = FALSE
)
```

The same plot can be restricted to a subset of IDs or a time range when there
are many individuals or many occasions.

```{r plot-subset, fig.width = 7, fig.height = 4}
PlotByID(
  data = data,
  id = id,
  time = time,
  observed = observed,
  ids = c(1, 2),
  times = c(0, 2),
  legend = TRUE,
  ask = FALSE
)
```

## 3. Remove rows with no finite observed values

`DeleteObservedAllNA()` is useful when rows with no observed measurements should
not contribute to elapsed-time construction or modeling.

```{r}
data_rows <- DeleteObservedAllNA(
  data = data,
  id = id,
  time = time,
  observed = observed,
  covariates = covariates
)

data_rows
```

## 4. Diagnose IDs before dropping anything

```{r}
diagnostics <- DiagnosticsByID(
  data = data_rows,
  id = id,
  time = time,
  observed = observed,
  covariates = covariates,
  min_nonmissing = 1L,
  extreme_cut = c(4, 5, 6),
  sd_cut = c(0.10, 0.05)
)

diagnostics
```

## 5. Convert diagnostics to flags

```{r}
flags <- FlagDiagnosticsByID(
  x = diagnostics,
  min_observed_rows = 2L,
  min_complete_rows = NULL,
  max_prop_all_missing = 0.95,
  max_gap = NULL,
  max_median_gap = NULL,
  min_sd = NULL,
  extreme_cut = 6,
  drop_score_cut = 2L
)

flags[flags$flag_any, c("id", "priority_score", "flag_reason")]
```

## 6. Choose sensitivity-analysis candidates

```{r}
drop_id <- GetDropID(
  x = flags,
  flagged_only = TRUE
)

drop_id
```

The returned IDs are candidates for exclusion or sensitivity analysis, not
automatic exclusions.

## 7. Optional detrending and scaling diagnostics

```{r}
scale_diagnostics <- DiagnoseScaleByID(
  data = data_rows,
  id = id,
  time = time,
  observed = observed,
  sd_min = 0.10,
  z_cut = 6,
  min_n = 2L,
  flagged_only = FALSE
)

scale_diagnostics
```

## 8. Optional transformations

After inspecting diagnostics, apply transformations explicitly.

```{r}
data_scaled <- ScaleByID(
  data = data_rows,
  id = id,
  time = time,
  observed = observed,
  covariates = NULL
)

data_scaled
```

The transformed data can also be plotted as a final check.

```{r plot-scaled, fig.width = 7, fig.height = 4}
PlotByID(
  data = data_scaled,
  id = id,
  time = time,
  observed = observed,
  legend = TRUE,
  ask = FALSE
)
```

This step-by-step workflow keeps the audit trail clear: diagnose first, decide
second, transform third.
